---
documentclass: sigplanconf
classoption:
  - preprint
  - numbers
header-includes:
  - \usepackage{amsmath}
  - \usepackage{graphicx}
  - \usepackage{float}
output:
  pdf_document:
    number_sections: true
    latex_engine: pdflatex
    keep_tex: true
    template: null  # needed to make the use of sigplan documentclass.
bibliography: bib.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

\title{Stretching Java™ HTTP Servers in the multi-core world}

\authorinfo{}
           {}
           {}
\maketitle

\begin{abstract}
Despite its importance, scalability of modern applications is poorly
understood. On the one hand, applications business-logic and workload keep
growing in complexity. On the other hand, fast-paced time-to-market and cloud
computing platforms have contributed to popularize the usage of languages
based on complex runtime systems, being Java \textsuperscript{TM} one of
the most used.

Because no two applications are alike or use the runtime in exactly the
same fashion, there is no guarantee that any single set of parameters will be
perfectly suited for every pair. This reality highlights how important it is
to conduct focused performance evaluation and tunning at runtime level. 

By running series of experiments, we found out that, with high statistical
confidence, the average throughput per-core of the aforementioned service
running on a virtual machine (VM) with 2 available cores is better than the
same service running on a VM with 4 cores. We are well aware that these
results ask for more investigation and answer those questions is definitely part of our future work.
\end{abstract}

```{r child = 'intro.rmd'}
```
```{r child = 'related.rmd'}
```
```{r child = 'exp.rmd'}
```
```{r child = 'results.rmd'}
```
```{r child = 'conc.rmd'}
```

# References {#references}